---
title: "Reappraisal of Tellez (2022) - Land Opportunism and Displacement in Civil Wars"
date:  "`r format(Sys.time(), '%a/%d/%b')`"
author: "46673"
always_allow_html: true
output: 
  bookdown::pdf_document2:
    toc: false
header-includes:
  - \usepackage{float}
---

```{r setup, include = FALSE}
# set global options 
knitr::opts_chunk$set(
  echo=FALSE, 
  warning=FALSE, 
  message=FALSE, 
  linewidth=60,
  fig.pos = "H",
  fig.align = "center"
  )

# library import:
source(file = "../code/package_manager.R")
update_geom_defaults("text", list(family = "sans"))

# other options:
options(scipen=999)

# run analysis scripts
source(file = "../code/analysis/replication-analysis.R")
source(file = "../code/analysis/callaway-analysis.R")
source(file = "../code/analysis/assumption-tests.R")
```

# 1  Introduction  {-}

The displacement of people from their homes is an ever-present consequence of 
war, as demonstrated by the tragic ongoing conflicts in Palestine, Sudan, and 
Ukraine: which have led to the forced displacement of many families over the 
course of the conflict. The resulting 
humanitarian crises warrant a focus on identifying the drivers of conflict-led
displacement, where identifying the causal mechanism that gives rise to 
displacement may shed light on how these humanitarian crises are likely to 
unfold. Tellez (2022) proposes to study the effect of opportunism as an 
underlying motivation for civilian displacement during a conflict, leveraging 
the introduction of palm plantations during the Colombian civil war as an 
indication of the potential for opportunistic displacement. They argue that the
expansion of the African palm oil industry in the 1990s led to an increase in 
the demand for land, resulting in the opportunistic acquisition of land via 
the forced displacement of civilians with the help of armed militant groups. To
this end, they implement a difference-in-difference design at the municipal-
year level utilising a panel dataset that includes an indicator variable for
the presence of palm oil plantations and the displacement rate in a given 
municipality during a given year.

They conclude that the years in which municipalities begin and continue the 
expansion of local palm oil industries record higher rates of civilian 
displacement compared to before they adopted the industry. They go on to model 
and conclude that this increased displacement is more pronounced in years where
palm-oil production is higher, and when a municipality has a historical record 
of paramilitary presence. In addition, they show that this relationship does not 
exist in municipalities with historical left-wing guerrilla presence, 
observing that opportunistic displacement is therefore reliant on the elite-
capture of the state, and not on violence alone.

While the author uses both municipal-level data and a household-level survey to
substantiate their claims, the difference-in-difference study is of particular
interest here as it requires a number of underlying assumptions to be true for 
its causal conclusions to hold water. I subject the author's research design to 
a battery of tests to investigate the validity of the conclusions they make, 
first reproducing select yet important parts of their analysis, and then 
providing evidence that the research design and analysis conducted by Tellez 
does not satisfy the necessary conditions to draw a clear causal relationship
between palm oil production and opportunistic displacement.

# 2  Research-Design and Background  {-}

The author uses a two-way fixed effects model (TWFE) to implement their diff-in-
diff design, regressing the inverse-sine-transformed displacement rate per
thousand residents on an indicator dummy variable recording the presence of a 
palm plantation in the municipality. As previously stated, the study is 
conducted at the municipality-year level. Further, they control for the number 
of left-wing guerrilla attacks and the presence of coca cultivation in the 
municipality in a given year, suggesting that guerrilla group behaviour and 
changes to the drug trade are likely to affect the costs of palm oil production
and thereby palm oil adoption within a municipality. These variables may affect 
displacement confounding the results: where municipalities with a higher 
guerrilla presence may have higher levels of conflict, resulting in a higher rate 
of forced displacement: and the presence of the drug trade in a municipality may 
induce opportunistic displacement due to coca cultivation.

$$
dispRate_{it} = \alpha_i + \omega_t + \beta \: palmOil_{it} + \phi \: X_{it} + \epsilon_{it}
$$
The adoption of palm oil within a municipality acts as the treatment assignment
where the treatment of a municipality is conditioned on coca cultivation, 
guerrilla presence, and the fixed effects are assumed to be as-if random 
resulting in a potential displacement rate that is independent of treatment. A 
further set of assumptions is required for the author's causal mechanism to 
hold, given that a TWFE model is applied to a dataset with staggered treatment 
timing. First, there should be no interference between municipalities, where in 
the case of the current research design, the treatment of any municipality 
should not spill over to another municipality. This assumption can be violated 
when palm oil adoption in a municipality causes palm oil adoption in its 
neighbouring municipalities, resulting in a spillover effect. Second, we require 
the conditional parallel trends assumption to hold, where the trend of the 
potential displacement rate for the treated municipalities if they were 
untreated should remain in parallel with the untreated municipalities' trend, 
conditional on the other covariates. Third, opportunists within municipalities 
must not anticipate the adoption of palm oil resulting in an early increase in 
the displacement rate, termed the no-anticipation assumption. Lastly, we require 
a homogeneous treatment effect over time. 

Once these assumptions are satisfied, the $\beta$ parameter identifies the TWFE
treatment effect estimand, allowing the author to conclude that palm oil 
adoption results in an increase in the displacement rate driven by opportunism, 
given that their transfer mechanism is justified. However, can we be sure these
assumptions are satisfied? While the author uses a Sun & Abraham and
synthetic control study to provide further evidence in favour of their 
conclusions, a more thorough set of tests are required to validate the causal 
conclusions made. Additionally, a static TWFE model is most appropriate when
there are no staggered treatment timings with heterogeneous treatment effects. 
Therefore, I propose a Callaway & Sant'Anna model which I implement and compare 
against the author's model choice to account for the characteristics of the 
research design.

# 3  Data  {-}

I download the raw municipality dataset from the author's online appendix 
repository, including any code used to replicate the results in R. The 
attached muni.rds file is briefly cleaned by reformatting some variables to 
make future exploratory and replicative analysis more convenient. I use the 
code provided by the author, making minor changes where necessary to replicate 
the analysis. Additionally, I access the Marco geoestadístico nacional MGN2005
shapefile datasets available on the Colombian National Administrative Department 
of Statistics website, providing me with fine-grained geospatial information 
on the municipalities with identification codes in line with the author's data.

The municipality data includes forced displacement counts provided by the 
Colombian Victim’s Unit, the palm oil indicator variable and palm oil production
values collected from the National Federation of Palm Oil Growers of Colombia. 
Alongside these variables, the author provides indicator functions that measure 
the historical presence of FARC (paramilitary) and AUC (guerrilla) groups used 
as controls in the study.

# 4  Replication of The Original Results  {-}

The results of the municipal-level analysis are replicated in Figure 1A, 2 
and Table 1 below. As Tellez discusses, Figure 1A suggests that the displacement 
rate is approximately similar between treated and control municipalities 
at the beginning of the 1990s. The rate subsequently diverges from the 2000s 
onward, indicating that the treatment effect may begin to take effect further 
into treatment. While the author makes an argument based on changes in the 
displacement rate, they proceed to conduct their diff-in-diff analysis using
the inverse-sine displacement rate. The same graph as 1A using the 
transformed rate is depicted in Figure 1B. First, the rate is distinctly 
different between both groups from the onset of the study. Second, the 
divergence in the rate between the two groups is not as pronounced after the 
transformation. These observations dampen the conclusion that Tellez makes where 
they state that "... Figure 1 below foreshadows some of the results of the 
analysis."

```{r displacement-plots, fig.cap="Displacement Rate for treated and control municipalities over time."}
f1 <- f1 + labs(title = "Figure 1A")
f2 <- f2 + labs(title = "Figure 1B")

shared_theme <- theme(
  plot.title.position = "plot", 
  plot.title           = element_text(size = 8, hjust = 0.5),
  plot.margin          = margin(10, 5, 5, 5)
)

combo <- (f1 + f2) +
  plot_layout(ncol = 2, widths = c(1,1)) &
  shared_theme & 
  theme(
    plot.background = element_rect(
      colour = "black",   
      size   = 0.5,       
      fill   = NA         
    ),
    plot.margin = margin(5,5,5,5)
  )

combo
```

The author goes on to construct four models, the results of which are depicted 
in Table 1 below. The results accurately depict a TWFE generalised diff-in-diff
design, modelling the effect of palm oil adoption on the inverse-sine transformed 
displacement rate. It is important to note that models 1, 2, and 3 only vary in 
terms of a single coefficient that interacts with the treatment variable of interest 
with the plantation treatment indicator. Given the model specification and 
underlying assumptions are accurate, the treatment effect under the study design
is positive, indicating that palm-oil adoption results in an increase in the 
forced displacement rate. Additionally, models 2 and 3 suggest that the increase
in the rate are driven both by increased production of palm-oil, and the 
presence of paramilitary groups. However, in both these models, the treatment 
effect coefficient flips sign, indicating that the baseline displacement rate
reduces with the adoption of a plantation, increasing only after a certain level 
of palm oil production or para-military presence is reached. Model 4 confirms 
the author's conclusion that the presence of guerrilla groups is not a 
significant determinant of displacement. Further, the marginal effect of palm
oil production and the presence of paramilitary groups are reproduced in 

Figure 2 below. 

\begin{table}[!htbp] \centering 
  \caption{Effect of palm-oil growth on displacement. Models include municipal 
  and year fixed effects and controls for time-varying presence of coca and FARC 
  attacks.} 
  \label{palm-pres-mods} 
\begin{tabular}{@{\extracolsep{5pt}}lcccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{4}{c}{\textit{Dependent variable:}} \\ 
\cline{2-5} 
\\[-1.8ex] & \multicolumn{4}{c}{Displacement rate (inverse-sine transformation)} \\ 
\\[-1.8ex] & (1) & (2) & (3) & (4)\\ 
\hline \\[-1.8ex] 
 Palm-oil plantation & 0.382$^{***}$ & $-$2.056$^{***}$ & $-$1.213$^{***}$ & 0.692 \\ 
  & (0.120) & (0.642) & (0.023) & (0.548) \\ 
  & & & & \\ 
 Plantation X Natl Prod (log) &  & 0.248$^{***}$ &  &  \\ 
  &  & (0.063) &  &  \\ 
  & & & & \\ 
 Plantation X AUC presence (dummy) &  &  & 1.614$^{***}$ &  \\ 
  &  &  & (0.120) &  \\ 
  & & & & \\ 
 Plantation X FARC presence (dummy) &  &  &  & $-$0.327 \\ 
  &  &  &  & (0.559) \\ 
  & & & & \\ 
\hline \\[-1.8ex] 
Observations & 14,192 & 14,192 & 14,192 & 14,208 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table} 

```{r marginal-plots, fig.cap="Marginal effect of palm oil production and para-military presence."}
p1 + p2 + 
  plot_layout(ncol = 2, widths = c(1,1)) & 
  theme(
    plot.background = element_rect(
      colour = "black",   
      size   = 0.5,       
      fill   = NA         
    ),
    plot.margin = margin(5,5,5,5)
  )
```

# 5  Alternative Estimation Using Callaway & Sant’Anna  {-}

I estimate the group-time ATT according to the Callaway & Sant'Anna study design
to account for the staggered treatment timing inherent in the data. I utilise a 
doubly-robust estimation procedure with 1000 bootstrap samples. The design has 
the added benefit of loosening the homogeneous treatment effect assumption 
required by a TWFE design. The group-time ATTs with the same group-time 
differences are averaged to compute the relative time ATT presented in Figure 3. 
Further, the simple average and weighted average ATT are presented in Table 2. 

```{r csa-plot, fig.cap="Relative-Time ATT based on the Callaway \\& Sant'Anna model."}
csa_plot + 
  theme(
    plot.background = element_rect(
      colour = "black",   
      size   = 0.5,       
      fill   = NA         
    ),
    plot.margin = margin(5,5,5,5)
  )
```

While the $\beta$ coefficient of the TWFE model presented a significantly 
positive relationship between palm oil adoption and displacement, both the 
simple and weighted average ATT under the Callaway & Sant'Anna model indicate
that there may be no overall effect, disputing the author's claims. The 
relative-time ATTs plotted in Figure 2 provide a more interesting picture, 
hinting at potential assumption violations that will require further tests to 
confirm. First, the post-treatment ATT remains around 0 for the first three 
years following treatment before rising to approximately 50% on average in the
fourth and fifth post-treatment years. This suggests that the treatment effect 
may begin to take effect later in treatment, similar to the author's conclusion
based on their plots in Figure 1 and appendix Figure A4. Second, the dip in the
pre-treatment ATT at the third and fourth years indicates that the inverse-sine 
displacement rate of soon-to-be treated municipalities defer from the control 
group municipalities, suggesting a potential violation of the parallel trends 
assumption. I test for these assumption violations next, as their results may
bring the authors' design, and thereby their conclusions, into question.

```{r csa-table}
df_table <- tibble::tribble(
  ~Measure,                               ~ATT,     ~SE,    ~`CI low`,  ~`CI high`,
  "Simple avg (equal‐wt across g,t)",     -0.0698,  0.361,   -0.777,    0.638,
  "Weighted avg (wt by group size)",      0.0519,  0.186,   -0.313,    0.417
)

df_table %>%
  kable(
    format    = "latex",
    booktabs  = TRUE,
    caption   = "ATT estimates: simple vs. weighted average",
    label     = "tab-att"
  ) %>%
  kable_styling(latex_options = c("hold_position"))
```

# 6  Diagnostic and Falsification Tests  {-}

I leverage four diagnostic and falsification tests to identify any potential 
violations of the key assumptions made in the two models above. First, the 
Goodman-Bacon decomposition is used to identify the determinants of the 
treatment effect estimated by the TWFE model. Once identified, we can reason as 
to how much of the effect results from comparing treated-vs-untreated 
municipalities. Second, a parallel trends test on the original TWFE model 
specification as well as a Sun and Abraham (2020) specification are utilised to 
provide multiple comparative viewpoints as to whether the parallel trends 
assumption is likely to be violated. Third, I test the no-interference 
assumption under SUTVA by testing for possible spillover effects from 
neighbouring treated municipalities: using geospatial data to identify 
neighbouring municipalities. Finally, I perform a test for no anticipation, 
using placebo treatment years in the past to test for any significant 
anticipatory ATT.

## 6.1  Goodman-Bacon Decomposition {-}

The Goodman-Bacon decomposition indicates that most of the identifying variation 
in the TWFE estimate comes from comparing municipalities that receive treatment 
with municipalities that are never treated, resulting in a positive treatment 
effect. Further, it implies that the bias from already-treated controls is 
small. Given that the TWFE estimate is an amalgamation of different 2x2 
comparisons, the decomposition results reflect positively on the 
interpretability of the estimate as a comparison of treated municipalities vs
untreated control municipalities. 

```{r bacon-table, results='asis'}
library(knitr)

bacon_table <- data.frame(
  Comparison      = c("Both Treated",
                      "Later vs Always Treated",
                      "Treated vs Untreated"),
  Weight          = c(0.00958, 0.04427, 0.94615),
  Avg_Estimate    = c(0.10943, 0.09587, 0.43840)
)

bacon_table %>%
  kable(
    format    = "latex",
    booktabs  = TRUE,
    caption   = "Goodman-Bacon decomposition of the TWFE estimate",
    col.names = c("Comparison", "Weight", "Average Estimate"),
    digits    = c(NA, 5, 5)
    ) %>%
  kable_styling(latex_options = c("hold_position"))
```

## 6.2  Parallel-Trends Test  {-}

The parallel trends assumption is tested using the relative-time ATTs from three 
different designs. Specifically, the Callaway and Sant'Anna specification in 
Figure 3 above and a dynamic TWFE, and Sun & Abraham specification, the results 
of which are depicted in Figure 4 below. Comparing the pre-treatment ATTs reveal
a shortcoming in the application of a TWFE model to a study with staggered 
treatment timing. The dynamic TWFE plot depicts a clear negative pre-treatment 
effect likely resulting from bias' given comparisons between "to be treated" and 
treated municipalities: comparisons that are dropped in the other designs 
resulting in next to no pre-treatment effect. Therefore, the parallel trends 
assumption is likely to hold given the pre-treatment effects in the Callaway and
Sun designs, corroborating Tellez's conclusions. However, I suggest some caution
when interpreting these results given the negative pre-treatment effect observed 
in the Callaway & Sant'Ana specification.

```{r pretrend-wald, fig.cap="Parallel trends test using a dynamic TWFE design and Sun and Abraham's design"}
ptrend_plot + 
  theme(
    plot.background = element_rect(
      colour = "black",   
      size   = 0.5,       
      fill   = NA         
    ),
    plot.margin = margin(5,5,5,5)
  )
```

## 6.3  Spillover (No-Interference) Test {-}

To conduct a spillover hypothesis test, I first use geospatial data obtained 
from the Colombian National Administrative Department of Statistics to identify 
neighboring municipalities, by selecting a municipality as a neighbor if the 
distance between the centroid of two municipalities is less than or equal to 
50 km. Once all neighbours are determined, I create a variable that records the 
proportion of treated neighbors in the previous year for each municipality-year 
record in the dataset. Given that the "did" package was used to implement the 
Callaway & Sant'Anna design does not currently implement a method to interact 
a spillover variable, I interact the spillover variable with each relative-time 
indicator in a dynamic TWFE model to estimate the spillover coefficients. 
Finally, a Wald hypothesis test is applied to the joint pre-treatment spillover 
coefficients to determine if a spillover effect is likely, the results of which
are displayed in Table 4 below. A p-value below 0.01 suggests that there is
strong evidence to reject the hypothesis that there is no spillover effect. This
should lead us to question if the no-interference assumption is likely to be 
satisfied, in this study design.

```{r spillover-test}
stat  <- wald_res$stat      
pval  <- wald_res$p        
df1   <- wald_res$df1          
df2   <- wald_res$df2

hyp <- "$H_0$: $\\beta_{relative\\_time=-10:spillover} = \\dots$ \\\\ $= 
\\beta_{relative\\_time=-2:spillover} = 0$"

test_tbl <- data.frame(
  Hypothesis = hyp,
  `Test Statistic`= stat,
  `P-value` = pval,
  `DF (num.)` = df1,
  `DF (den.)` = df2,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

test_tbl %>% kable(
  format = "latex",
  booktabs = TRUE,
  caption = "Wald test for the joint nullity of the pre-treatment spillover coefficients",
  escape = FALSE,
  digits = c(3)
) %>%
    kable_styling(latex_options = c("hold_position"))
```

## 6.4  No-Anticipation Placebo Test {-}

Finally, I subtract 2 years and 3 years from the treatment year 
variables, dropping any treatment years that fall below the minimum year of the 
study to create a placebo treatment indicator. This placebo is used as the 
treatment variable in two Callaway & Sant'Anna models, the results of which are
depicted in Figures 5A and 5B, respectively. The post-treatment ATTs in the 2-year
model hover close to or just below zero, which suggests that the no-anticipation
assumption is likely to hold. However, the 3-year model identifies a negative 
ATT across all post-treatment years. While this may suggest a violation of the 
no-anticipation assumption; it may be a potential artefact of the negative 
pre-treatment ATT observed in the parallel trends test. Altogether, some caution 
is advised in the interpretation of all model coefficients, given the anomalous 
observations in both tests.

```{r placebo-test, fig.cap = "Callaway \\& Sant'Anna placebo (no-anticipation) test."}
placebo_plot_1  <- placebo_plot_1 + 
  labs(title = "Figure 5A - placebo treatment 2 years before")
placebo_plot_2  <- placebo_plot_2 + 
  labs(title = "Figure 5B - placebo treatment 3 years before")

shared_theme <- theme(
  plot.title.position = "plot", 
  plot.title = element_text(size = 8, hjust = 0.5),
  plot.margin = margin(10, 5, 5, 5)
)

placebo_plot_1 + placebo_plot_2 +
  plot_layout(ncol = 2, widths = c(1,1)) &
  shared_theme & 
  theme(
    plot.background = element_rect(
      colour = "black",   
      size = 0.5,       
      fill = NA         
    ),
    plot.margin = margin(5,5,5,5)
  )
```

# 7  Conclusion  {-}

While a TWFE model does not always result in a convenient interpretation of the 
ATT due to the different components that make up the estimand given a staggered
treatment timing, the Goodman-Bacon decomposition does provide some solace in 
that the comparison being made is between the treated and the untreated 
resulting in a believable Average Treatment Effect that is not exposed to any
negatively weighted components. However, the ability for the specification to 
identify this ATT depends on a series of strict assumptions. The tests conducted 
provide evidence to reject the no-interference assumption and cast some doubt 
on the validity of the parallel trends and no-anticipation assumption: although 
the jury is not yet out on this front. Therefore, I suggest some caution when 
interpreting the author's central claim, that the adoption of palm oil within 
a municipality results in an increase in forced displacement driven by an
opportunistic elite-state interaction. 

I provide an alternative model (Callaway & Sant'Anna), alongside the generalised 
TWFE model and the Sun and Abraham model included in the author's online 
appendix, that make valid comparisons between treated and untreated 
municipalities providing further support in favour of the author's claim that the
ATT is significant and positive at some post-treatment years. However, it is 
important that future work include additional controls that may improve
the validity of making a conditional parallel trends assumption. Alternatively, 
the parallel trends assumption may hold when using the raw displacement rate or 
log displacement rate, given that the parallel trends assumption is sensitive to 
the functional form of the response variable. Additionally, I suggest purging 
the spillover effect before model estimation which may result in a more
believable group-time ATT in the case of the Callaway & Sant'Anna model. Modern 
estimation techniques do result in variable point estimates due to an issue of 
low power. Therefore, I recommend revisiting the study design given the 
mounting evidence in favour of several violations of the causal assumptions.

------------------------------------------------------------------------

\clearpage

# Code appendix

```{r ref.label = knitr::all_labels(), echo=TRUE, eval=FALSE}
```
